---
title: "Handling difficult data with R 1: Air Quality data"
output: html_notebook
---

## Introduction

This is a notebook exploring a difficult data set that was given to me in a job exam - that I failed. I attempted the exam in python and well it didn't go well. This is the first of a series of notebooks that explores data processing and wrangling in R. [There is a complementary notebook for this project as well](../AirQuality_dealing_with difficult_data.ipynb)

## 1. Data Loading and tidying

```{r}
library(tidyverse)
```


```{r,eval=False}
df <- read.csv("../AirQualityUCI.csv")
```
Already we have an issue loading the data

```{r}
df <- read.csv("../AirQualityUCI.csv",sep=';')
head(df)
```
The issue was the same as in Python and pandas, I needed to specify the separator, which is a semi-colon as opposed to a comma. As with python, we see there are 2 additional columns X and X.1

```{r}
colnames(df)
```

Remove X and X.1

```{r}
df$X.1 <- NULL
df$X <- NULL

colnames(df)
```


Missing/null values


```{r}
df %>% is.na() %>% colSums()
```


```{r}
dim(df)
```

```{r}
df[is.na(df),]
```

912 rows? It was only 114 rows! WTF? Maybe `complete_cases` will work?


```{r}
df[!complete.cases(df),]
```

Seems more sensible. Remove these data 

```{r}
df <- df[complete.cases(df),]
dim(df)
```

Good, now some basic info

```{r}
str(df)
```

```{r}
summary(df)
```

There are a number of issues with this data set that needs to be fixed:
  1. 5 columns are factors when they should be floats: CO.GT., C6H6.GT., T, RH, AH  
  2. minimum values of -200 should be replaced with Na and imputed
  
First let's deal with the 5 columns


```{r}
df %>% select(CO.GT.) %>% head()
```

As with the Python problem we need to replace the comma with a period and convert to float

```{r}
df[1:10,"CO.GT."] %>% str_replace(.,",",".") %>% as.numeric()
```

Good now to do for all

```{r}
cols_ <- c("CO.GT.", "C6H6.GT.", "T", "RH", "AH" )
for (col in cols_) {
  #df[,col] <- df %>% select(col) %>% str_replace(.,",",".") %>% as.numeric()
  df[,col] <- df[,col] %>% as.character() %>% str_replace(.,",",".")  %>% as.numeric()
}

str(df)
```


The next stage is to replace the -200 values with NA. let's take a sample data. Will use the `na_if` function from `dplyr`/`tidyverse`

```{r}
temp <- df[1:20,"CO.GT."]
temp[18] <- -200
temp
```

```{r}
(temp_na <- na_if(temp, -200))
```

Good now to do for all columns



```{r}
for (col in colnames(df)) {
  #df[,col] <- df %>% select(col) %>% str_replace(.,",",".") %>% as.numeric()
  df[,col] <- na_if(df[,col], -200)
}

df %>% is.na() %>% colSums()
```



```{r}
df %>% is.na() %>% colMeans()
```

Good the same results as with Python

```{r}
summary(df)
```


Good the data seems to be in a format that is most reasonable. The next step is some basic data exploration that will require some manipulation of the data.

### 2. Data Exploration 

NMHC(GT) is missing 90% of data! That is significant, and most probable will have to be dropped. Before moving to the next stage (imputing and model training), let's answer some questions regarding the data:

1. What is the highest CO content in June 2004? How does this compare with December 2004?
2. The most common time and weekday of the top 10 highest NO2, CO and T. 
3. The month with the highest average CO? 
4. The 5 features most correlate with temperature and absolute humidity? Are they the same? 

 
#### 1.  What is the highest CO content in June 2004? How does this compare with December 2004?

before continuing we need to convert the Date and Time columns to a datetime format


```{r}
df %>% select(Date,Time) %>%head()
```

use the `anytime` package

```{r}
anytime::anytime(df$Date[1])
```

Ok this is a bit annoying. The timeformat is wrong. Looking online I need to look at the formats

```{r}
anytime::getFormats()
```

```{r}
"%d/%m/%Y" %in% anytime::getFormats()
```

This format does not exist, so we should add it as well as a few more


```{r}
anytime::addFormats(c("%d/%m/%Y","%d/%m/%Y %H:%M:%S", "%d/%m/%Y %H.%M.%S"))
anytime::anytime(df$Date[1])
```

huzza! Let's try with the Time format

```{r}
anytime::anytime(paste0(df$Date[1], ' ' ,df$Time[1]))
```


Great! ow we can add another column

```{r}
df$Timestamp <- anytime::anytime(paste0(df$Date, ' ' ,df$Time))
df %>% select(Date, Time, Timestamp)
```
Noice! Now we can proceed with the question

max in June

```{r}
# add a month column

df$month <- lubridate::month(df$Timestamp)

df %>% filter(month == 6) %>% select(CO.GT.) %>% summarise(max(CO.GT.,na.rm = TRUE))

```

 Max in December
```{r}
df %>% filter(month == 12) %>% select(CO.GT.) %>% summarise(max(CO.GT.,na.rm = TRUE))
```

Bonus: max per month

```{r}
df %>% group_by(month) %>% select(CO.GT.) %>% summarise(max(CO.GT.,na.rm = TRUE))
```

The highest month is November



#### 2. The most common time and weekday of the top 10 highest NO2, CO and T. 


Add a new colun with Weekday


```{r}

df$weekday <- weekdays(df$Timestamp)
```


```{r}
df %>% select(Time,weekday, NO2.GT.) %>% arrange(desc(NO2.GT.)) %>% head(10)
```

The most common time is 09:00:00 and on Friday

```{r}
df %>% select(Time,weekday, CO.GT.) %>% arrange(desc(CO.GT.)) %>% head(10)
```
 The most common time is 18:00:00 and day Tuesday



```{r}
df %>% select(Timestamp,weekday, T) %>% arrange(desc(T)) %>% head(10)
```

The most common time is 16:00:00 and day is Thursday

#### 3. The month with the highest average CO?

```{r}
df %>% 
  select(month, CO.GT.) %>%
  group_by(month) %>% 
  summarise(
    mean_CO = mean(CO.GT.,na.rm=TRUE)
    ) %>% 
  arrange(desc(mean_CO))
```


The month with the highest average CO is December



#### 4. The 5 features most correlate with temperature and absolute humidity? Are they the same? 


```{r}

cor(df[,colnames(df)[3:15]], use="complete.obs")[,"T"]
```

```{r}
cor(df[,colnames(df)[3:15]], use="complete.obs")[,"AH"]
```


### 3. Building a predictive model



